Simulation Learning in Computer Hex

### 2011-11-28 thesis/thesis.tex
Cleaned up language and grammar.

\section{Introduction}
Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. The choice of game and rules can also be experimentally chosen for a variety of difficulty and computational complexity levels. Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World-Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. For at least Checkers, Chess, and similar games (Othello, Connect Four), brute force tree search has provided a silver bullet.

@@

\section{Background: The Game of Hex}
Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons. The players are denoted by colors here \black{} and \white. The object of the game is to form an unbroken chain from one side of the board to the other. Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board. Figure \ref{fig:13x13finished} shows the final position, \black{} having just played the winning move, with an unbroken chain connecting both his sides. Hex has properties that make it attractive for AI research. Once placed, a piece never leaves the board. The only criterion for a legal move is that the hexagon be unoccupied (empty). Gale '79 proved the game cannot end in a draw, simplifying minimax methods. Nash, one of the inventors of the game, used a strategy stealing argument to prove the existence of a winning strategy for the first player, but the proof is by reduction and offers no insight into the strategy itself. Solutions for all opening positions for 7x7 and 8x8 boards have been found\cite{henderson2009solving}. A good portion, but not all, 9x9 openings have also been solved. Nash proposed 11x11 as the ideal board size, but 13x13 has been more common in competitions, especially as the ability of computer players improves.

@@

\subsection{Update Policy}
Once a simulation is finished, the result is used to update the tree statistics. In the basic implementation, only the win and visit statistics along the traversed path are updated. It was quickly discovered that more sophisticated update strategies could be used to keep alternate statistics, and update them at nodes similar to those used in the playout that were not in the path. One simple example of this is the All-Moves-As-First (AMAF) heuristic. Briefly, it leverages the fact that moves played in the simulation could have been played in any order. Thus, a move deep in the simulation could be fictitiously referred to as the ``first'' move. In addition to updating the statistics of the nodes in the path, all siblings corresponding to moves made in the simulation are also updated with a fictitious AMAF win and visit. Variations on AMAF have been tested, including Permutations AMAF, which updates node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.

@@

\subsection{Biasing the Simulation}\label{bias}
Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial\cite{gelly2006modification}\cite{gelly2008achieving}. However, this can require expert knowledge (heuristics) that is hard to obtain or not available for some games (particularly new games). They also found found that improving the performance of playout policy did not guarantee an improvement in the tree search. For example, the playout policy can implicitly prune good moves, never allowing them to be played. In Go, the policy must correctly handle complicated situations like nakade (life and death). If it fails to correctly play out a complex position, the tree might fail to see an opponent response. Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situation. If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive amount of playouts to ``find'' the move sequence. In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in the selection policy\cite{gelly2007combining}. This puts most of the control of the search in the hands of the simulation policy.

@@

\subsection{The Hivemind System}
Hivemind was designed for experiments in learning to play abstract strategy games on a regular board. The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``game tracker interface''. The design was influenced by Fuego, although Hivemind is written in a different language (Go). In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games\cite{Fuego}. Hivemind's tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe. The major contribution of Hivemind is the integration of evolutionary learning into the search process. Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.

\subsubsection{Learning}
The learning module is responsible for driving the evolutionary search process. Currently, Hivemind uses Evolution Strategies (see Section \ref{estrats}), although previous versions experimented with Particle Swarm Optimization. As described earlier, the learning process consists of three steps, iterated once per generation: selection, mutation, and evaluation. In general, the learning module is responsible for finding a strategy (described in Section \ref{esmods}) that generates optimal results during the evaluation step. Evaluation is performed by selecting two individuals from the population and having them play a game against each other, each individual parameterized by their strategy. The individual that wins receives points. During the selection step, individuals are culled based on points received, and the process continues. To play a game, the learning module iteratively calls into the search module with the strategy, until the tracker module reports the game is over. The tracker module is also used to collect the game results.

@@

\subsubsection{Game Tracker}

\section{Results and Conclusions}\label{results}
To test the effects of learning policy weights, the evolution process was ran for 41 generations. This took approximately 2 days using a single core of an Intel Xeon CPU running at 2.80GHz. After training was completed, the best individual from the 41st generation was selected for comparison. This individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally. The control consisted of uniform random playouts. All players used a slightly modified UCT selection policy that included the variance term suggested by Gelly 2006\cite{gelly2006exploration}. A pure update policy was used, only the nodes directly in the path traversed by UCT were updated with the result of a simulation.

200 games per pair were played, with time controls set at 5 seconds per move. All players have approximately the same performance with regard to simulations per second, so the 5 seconds per move resulted in around 10,000 simulated playouts. A boardsize of 13x13 was used, and to control for the swap rule a swap-safe move was used as the first move for black. Matches were run using the Gomill software\cite{gomill}. Gomill required no modification to work with Hex playing programs.

\begin{figure}
	\begin{center}
		\begin{tabular}{c | c c c}
		 & default & uniform & uniform (tenuki) \\
		\hline
		uniform & 63.50\% & & \\
		uniform (tenuki) & 64.00\% & 50.00\% & \\
		learned & 79.50\% & 81.50\% & 76.00\% \\
		\end{tabular}
	\end{center}
	\caption{All-play-all tournament of the 4 Hivemind variants.}
	\label{fig:results}
\end{figure}

Learned local play is the clear winner. Interestingly, the addition of a tenuki probability lowers performance in all cases. As expected, uniform local play significantly improves upon pure random playouts. The fact that evolutionary learning can find weights that boost performance versus uniform local in the same manner that uniform local improves upon pure random shows how much further there is to go, and shows the weaknesses of purely random simulations.

As an outside control, MoHex from the University of Alberta was compiled and ran with the default settings\cite{mohex}. Again, 200 games were played. As MoHex is a particularly strong competitor, Hivemind used 100,000 simulations per move. Default time settings for MoHex are 10 seconds, but MoHex uses some time to perform game theoretic search and pruning before the Monte-Carlo phase, and will stop the search early if it believes no better moves will be found. The pre-search can also discover ``must-play'' moves that will be immediately played in lieu of the MCTS. As a result, the time controls are not directly comparable. However, CPU time was tracked and can be used for a cost comparison.

\begin{figure}
	\begin{center}
		\begin{tabular}{c c c c}
		Player 1 & Player 2 & Win-rate & Relative CPU \\
		\hline
		Hivemind Default & MoHex & 6.5\% & 664\% \\
		Hivemind Learned & MoHex & 0\% & 555\% \\
		Hivemind Learned & Hivemind Default & 92.5\% & 131\% \\
		\end{tabular}
	\end{center}
	\caption{Win-rate and relative CPU used of Hivemind versus MoHex. Relative CPU is the ratio of average CPU time per move for Player 1 vs. Player2.}
	\label{fig:relativecpu}
\end{figure}

Learned local patterns provided the only wins against MoHex, with 13 wins out of 200 played. Despite the low performance, MoHex used about 1/6th to CPU time, showing a large distance to cover to get Hivemind up to competition strength. In this series, Hivemind used 100,000 playouts per move, approximately 10 times more than the self-play results. This resulted in an increase in self-play strength, winning 92.5\% of the time when using learned patterns, up from 81.5\% with 10,000 moves. The relative cost for this performance increase was very reasonable, with pattern computation requiring a 31\% overhead.

>>>>

\section{Introduction}
Abstract strategy games provide an attractive testbed for artificial intelligence and machine learning applications. The game framework provides immediate feedback on algorithm performance. The choice of game and rules can also be experimentally chosen for a variety of difficulty and computational complexity levels. Starting with mathematical game theory research, the most common technique for game AI has been brute force search of the tree of legal moves. This technique, highly tuned and optimized, is what led to Chinook's ascendance to World Champion in 1994 and Deep Blue's eventual victory in 1997 over then World Champion Gary Kasparov. These victories cemented brute force tree search as standard techniques in abstract strategy games, to the point where computers constructing endgame tables have been strong enough to find sequences lasting for hundreds of moves. Kasparov himself suggested a version of Chess that allowed the player to consult a computer for move advice. Chinook went so far as to solve checkers, proving that perfect play will end in a draw. For at least Checkers, Chess, and similar games (Othello, Connect Four), brute force tree search has provided a silver bullet.

@@

\section{Background: The Game of Hex}
Hex is a two player, full-information, abstract strategy game played on an $N$x$N$ grid of hexagons. The players are denoted by  a color, either \black{} and \white. The object of the game is to form an unbroken chain from one side of the board to the other. Figure \ref{fig:13x13inprogress} shows an example game in progress on a 13x13 board. Figure \ref{fig:13x13finished} shows the final position, \black{} having just played the winning move, with an unbroken chain connecting both his sides. Hex has properties that make it attractive for AI research. Once placed, a piece never leaves the board. The only criterion for a legal move is that the hexagon be unoccupied (empty). Gale '79 proved the game cannot end in a draw, simplifying minimax methods. Nash, one of the inventors of the game, used a strategy stealing argument to prove the existence of a winning strategy for the first player, but the proof is by reduction and offers no insight into the strategy itself. Solutions for all opening positions for 7x7 and 8x8 boards have been found\cite{henderson2009solving}. A good portion, but not all, 9x9 openings have also been solved. Nash proposed 11x11 as the ideal board size, but 13x13 has been more common in competitions, especially as the ability of computer players improves.

@@

\subsection{Update Policy}
Once a simulation is finished, the result is used to update the tree statistics. In the basic implementation, only the win and visit statistics along the traversed path are updated. It was quickly discovered that more sophisticated update strategies could be used to keep alternate statistics, and update them at nodes similar to those used in the playout that were not in the path. One simple example of this is the All-Moves-As-First (AMAF) heuristic. Briefly, it leverages the fact that moves played in the simulation could have been played in any order. Thus, a move deep in the simulation could be fictitiously referred to as the ``first'' move. In addition to updating the statistics of the nodes in the path, all siblings corresponding to moves made in the simulation are also updated with a fictitious AMAF win and visit. Variations on AMAF have been tested, including Permutations AMAF, which updates a node if any permutation of the simulation contained the node's position, and Some-Moves-As-First, which attempts to restrict the number of siblings updated.

@@

\subsection{Biasing the Simulation}\label{bias}
Gelly and others have shown that using heuristics in both the expansion and playout policy can prove extremely beneficial\cite{gelly2006modification}\cite{gelly2008achieving}. However, this can require expert knowledge (heuristics) that is hard to obtain or not available for some games (particularly new games). They also found found that improving the performance of playout policy did not guarantee an improvement in the tree search. For example, the playout policy can implicitly prune good moves, never allowing them to be played. In Go, the policy must correctly handle complicated situations like nakade (life and death). If it fails to correctly play out a complex position, the tree might fail to see an opponent response. Much time and energy has been spent on hand-tuning the policies of the top engines to correctly handle complicated tactical situations. If there is some small probability of playing a purely random move, the convergence properties of the UCT algorithm will still hold, but can require a prohibitive amount of playouts to ``find'' the move sequence. In fact, one of the top Go engines, MoGo, uses a near-zero exploration coefficient in the selection policy\cite{gelly2007combining}. This puts most of the control of the search in the hands of the simulation policy.

@@

\subsection{The Hivemind System}
Hivemind was designed for experiments in learning to play abstract strategy games on a regular board. The system consists of three modules: the evolutionary learning process, a Monte-Carlo based UCT search tree implementation, and a ``game tracker interface'' used by the tree search to maintain game state information. The design was influenced by Fuego, although Hivemind is written in a different language (Google's Go\cite{golang}). In particular, the tracker interface is very similar to Fuego's subclassing system for implementing multiple games\cite{Fuego}. Hivemind's tracker interface is currently implemented for the games of Hex, Go, and Tic-Tac-Toe. The major contribution of Hivemind is the integration of evolutionary learning into the search process. Hivemind is also open-source software, available at \url{http://github.com/etherealmachine/hivemind}.

\subsubsection{Learning}\label{learning}
The learning module is responsible for driving the evolutionary search process. In general, the learning module is responsible for finding a strategy (described in Section \ref{esmods}) that performs well during the evaluation step. Hivemind currently uses Evolution Strategies, although previous versions experimented with Particle Swarm Optimization. Selection and mutation are implemented as described in section \ref{estrats}. Evaluation is described in section \ref{fitness}.

@@

\subsubsection{Game Tracker}
The game tracker allows complete games to be played vi Monte-Carlo Tree search. The implementation of a game has four responsibilities:
\begin{enumerate}
\item{Maintain all required game state. A single function, \texttt{Play(color, vertex)}, is used by the search tree to apply a move for the given color.}
\item{Report legal moves from a given game state. The \texttt{Legal(color, vertex)} function is used to query the legality of a given move. The search tree uses this to expand a parent node, adding all legal moves as children.}
\item{Report on the status of the winner. The \texttt{Winner()} function is used to determine the current winner of the game. The special \texttt{Empty} player is returned while the game is in progress, and either \black{} or \white{} once a winner has been determined.}
\item{Playing out the game from the current game state. The tracker is parameterized with the policy weights, and responsible for using them during the playout. The speed of the playout function is also critical, as more playouts-per-second generally result in better move selection. This emphasis on speed is why the tracker is required to implement a monolithic \texttt{Playout} function. Many games have optimizations that can be performed in the playout that the search tree is unaware of, for example, the Hex tracker keeps a simple list of randomized empty positions, so random move selection is a O(1) operation. This is because in Hex, any empty position is legal. This is not the case for Go, so the Go engine uses other optimizations for fast random move selection.}
\end{enumerate}
The additional \texttt{WasPlayed(color, vertex)} function provides information required for the AMAF heuristic. A few extra functions also provide for debugging by pretty-printing the tracker's internal state.

The game tracker interface provides a simple definition of what is needed for a game to use the MCTS+UCT framework. It also constrains the space of games that can be easily implemented in Hivemind. In particular, Hivemind allows for 2 player games with perfect information. The game space is not required to be a regular board. Node expansion uses the \texttt{Legal(color, vertex)} function to find all legal children. Vertices are assumed to be numbered consecutively, so a tracker for a non-regular board can encode vertices as it wishes.

\section{Results and Conclusions}\label{results}
To test the effects of learning policy weights, the evolution process was run for 100 generations. The Hivemind learning parameters used were: \texttt{generations=100, lambda=40, mu=30, parents=2, samples=5}. See sections \ref{estrats} and \ref{learning} for a description of the learning parameters. Each game was played with boardsize of 7x7 and 1000 playouts per move. After 100 generations were complete, the best individual from the 100th generation was selected for comparison. This individual was used to bias the simulation policy and compared to the uniform local bias, uniform local bias with a $\frac{5}{6}$th probability of playing locally. The control consisted of uniform random playouts. All players used a slightly modified UCT selection policy that included the variance term suggested by Gelly 2006\cite{gelly2006exploration}. A pure update policy was used, only the nodes directly in the path traversed by UCT were updated with the result of a simulation.

200 games per pair were played, with time controls set at 5 seconds per move. All players have approximately the same performance with regard to simulations per second, so the 5 seconds per move resulted in around 10,000 simulated playouts. A boardsize of 7x7 was used, and to control for the swap rule a swap-safe move was used as the first move for black. Matches were run using the Gomill software\cite{gomill}. Gomill required no modification to work with Hex playing programs. Results are shown in figure \ref{fig:results}.

\begin{figure}
	\begin{center}
		\begin{tabular}{c | c c c}
		 & default & uniform & uniform (tenuki) \\
		\hline
		uniform & 70.50\% & & \\
		uniform (tenuki) & 61.00\% & 50.00\% & \\
		learned & 90.00\% & 84.00\% & 86.00\% \\
		\end{tabular}
	\caption{All-play-all tournament of the 4 Hivemind variants. Each element is the percent win-rate of the row variant versus the column variant.}
	\label{fig:results}
	\end{center}
\end{figure}

Learned local play is the clear winner. Interestingly, the addition of a tenuki probability has no or little effect on performance in all cases. As expected, uniform local play significantly improves upon pure random playouts. The fact that evolutionary learning can find weights that boost performance versus uniform local in the same manner that uniform local improves upon pure random shows how much further there is to go, and shows the weaknesses of purely random simulations.

As an outside control, MoHex from the University of Alberta was compiled and run with the default settings\cite{mohex}. Again, 200 games were played. As MoHex is a particularly strong competitor, Hivemind used 100,000 simulations per move. MoHex played with the default time settings of 10 seconds per move, but it uses some time to perform game theoretic search and pruning before the Monte-Carlo phase, and will stop the search early if it believes no better moves will be found. The pre-search can also discover ``must-play'' moves that will be immediately played in lieu of the MCTS. As a result, the time controls are not directly comparable. However, CPU time was tracked and can be used for a cost comparison. Figure \ref{fig:relativecpu} shows the results of this series.

\begin{figure}
	\begin{center}
		\begin{tabular}{c c c c}
		Player 1 & Player 2 & Win-rate & Relative CPU \\
		\hline
		Hivemind Default & MoHex & 0\% & 555\% \\
		Hivemind Learned & MoHex & 6.5\% & 664\% \\
		Hivemind Learned & Hivemind Default & 92.5\% & 131\% \\
		\end{tabular}
		\caption{Win-rate and relative CPU used of Hivemind versus MoHex. Relative CPU is the ratio of average CPU time per move for Player 1 vs. Player2.}
		\label{fig:relativecpu}
	\end{center}
\end{figure}

Learned local patterns provided the only wins against MoHex, with 13 wins out of 200 played. MoHex strongly outperformed Hivemind, but used only a fraction (about 1/6th) of the CPU time, showing a large distance to cover to get Hivemind up to competition strength. In this series, Hivemind used 100,000 playouts per move, approximately 10 times more than the self-play results. This resulted in an increase in self-play strength, winning 92.5\% of the time when using learned patterns, up from 90\% with 10,000 moves. The relative cost for this performance increase was very reasonable, with pattern computation requiring a 31\% overhead.
